---
title: "Motorcycle Price Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Motorcycle Price Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/nehalbirla/motorcycle-dataset\>*

### Reference:

*\<Birla, N. (n.d.). Motorcycle Dataset [Data set]. Kaggle. https://www.kaggle.com/datasets/nehalbirla/motorcycle-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset
```{r Load dataset}
# Load dataset with specified column classes
vehicle_data <- read.csv("BIKE_DETAILS.csv", colClasses = c(
  name = "character",
  year = "integer",
  seller_type = "factor",
  owner = "factor",
  km_driven = "integer",
  ex_showroom_price = "numeric",
  selling_price = "numeric"
))

# Verify the structure to ensure column types are correct
str(vehicle_data)

# Display the first few rows to confirm data loading
head(vehicle_data)
View(vehicle_data)
```

## Measures of Frequency
```{r MOF}
# Load necessary library for summary statistics
library(dplyr)

# Frequency of seller type
seller_type_freq <- table(vehicle_data$seller_type)
print(seller_type_freq)

# Frequency of owner type
owner_freq <- table(vehicle_data$owner)
print(owner_freq)
```

## Measures of Central Tendency
```{r MOCT}
# Mean, median, and mode for numerical columns

# Mean
mean_year <- mean(vehicle_data$year, na.rm = TRUE)
mean_km_driven <- mean(vehicle_data$km_driven, na.rm = TRUE)
mean_ex_showroom_price <- mean(vehicle_data$ex_showroom_price, na.rm = TRUE)
mean_selling_price <- mean(vehicle_data$selling_price, na.rm = TRUE)

# Median
median_year <- median(vehicle_data$year, na.rm = TRUE)
median_km_driven <- median(vehicle_data$km_driven, na.rm = TRUE)
median_ex_showroom_price <- median(vehicle_data$ex_showroom_price, na.rm = TRUE)
median_selling_price <- median(vehicle_data$selling_price, na.rm = TRUE)

# Mode function
get_mode <- function(v) {
  uniq_v <- unique(v)
  uniq_v[which.max(tabulate(match(v, uniq_v)))]
}

mode_owner <- get_mode(vehicle_data$owner)
mode_seller_type <- get_mode(vehicle_data$seller_type)

print(list(
  mean_year = mean_year, mean_km_driven = mean_km_driven,
  mean_ex_showroom_price = mean_ex_showroom_price, mean_selling_price = mean_selling_price,
  median_year = median_year, median_km_driven = median_km_driven,
  median_ex_showroom_price = median_ex_showroom_price, median_selling_price = median_selling_price,
  mode_owner = mode_owner, mode_seller_type = mode_seller_type
))

```

## Measures of Distribution
```{r MOD}
# Variance and standard deviation
variance_km_driven <- var(vehicle_data$km_driven, na.rm = TRUE)
sd_km_driven <- sd(vehicle_data$km_driven, na.rm = TRUE)

variance_ex_showroom_price <- var(vehicle_data$ex_showroom_price, na.rm = TRUE)
sd_ex_showroom_price <- sd(vehicle_data$ex_showroom_price, na.rm = TRUE)

variance_selling_price <- var(vehicle_data$selling_price, na.rm = TRUE)
sd_selling_price <- sd(vehicle_data$selling_price, na.rm = TRUE)

# Range
range_year <- range(vehicle_data$year, na.rm = TRUE)
range_km_driven <- range(vehicle_data$km_driven, na.rm = TRUE)
range_ex_showroom_price <- range(vehicle_data$ex_showroom_price, na.rm = TRUE)
range_selling_price <- range(vehicle_data$selling_price, na.rm = TRUE)

# Range
range_year <- range(vehicle_data$year, na.rm = TRUE)
range_km_driven <- range(vehicle_data$km_driven, na.rm = TRUE)
range_ex_showroom_price <- range(vehicle_data$ex_showroom_price, na.rm = TRUE)
range_selling_price <- range(vehicle_data$selling_price, na.rm = TRUE)

# Quantiles
quantiles_km_driven <- quantile(vehicle_data$km_driven, na.rm = TRUE)
quantiles_ex_showroom_price <- quantile(vehicle_data$ex_showroom_price, na.rm = TRUE)
quantiles_selling_price <- quantile(vehicle_data$selling_price, na.rm = TRUE)

# Interquartile range (IQR)
iqr_km_driven <- IQR(vehicle_data$km_driven, na.rm = TRUE)
iqr_ex_showroom_price <- IQR(vehicle_data$ex_showroom_price, na.rm = TRUE)
iqr_selling_price <- IQR(vehicle_data$selling_price, na.rm = TRUE)

print(list(
  variance_km_driven = variance_km_driven, sd_km_driven = sd_km_driven,
  variance_ex_showroom_price = variance_ex_showroom_price, sd_ex_showroom_price = sd_ex_showroom_price,
  variance_selling_price = variance_selling_price, sd_selling_price = sd_selling_price,
  range_year = range_year, range_km_driven = range_km_driven,
  range_ex_showroom_price = range_ex_showroom_price, range_selling_price = range_selling_price,
  quantiles_km_driven = quantiles_km_driven, quantiles_ex_showroom_price = quantiles_ex_showroom_price,
  quantiles_selling_price = quantiles_selling_price,
  iqr_km_driven = iqr_km_driven, iqr_ex_showroom_price = iqr_ex_showroom_price,
  iqr_selling_price = iqr_selling_price
))

```

## Measures of Relationship
```{r MOR}
# Correlation matrix for numerical variables
cor_matrix <- cor(vehicle_data %>% select(year, km_driven, ex_showroom_price, selling_price), use = "complete.obs")
print(cor_matrix)

# Covariance matrix
cov_matrix <- cov(vehicle_data %>% select(year, km_driven, ex_showroom_price, selling_price), use = "complete.obs")
print(cov_matrix)
```

## ANOVA
```{r ANOVA}
## ANOVA test for selling price across different seller types
anova_seller_type <- aov(selling_price ~ seller_type, data = vehicle_data)
summary(anova_seller_type)

# ANOVA test for selling price across different ownership categories
anova_owner <- aov(selling_price ~ owner, data = vehicle_data)
summary(anova_owner)

# Perform Tukey's HSD post-hoc test if ANOVA is significant
tukey_seller_type <- TukeyHSD(anova_seller_type)
print(tukey_seller_type)

tukey_owner <- TukeyHSD(anova_owner)
print(tukey_owner)
```

## Plots
```{r Plots}
library(ggplot2)

# Histogram of selling prices
ggplot(vehicle_data, aes(x = selling_price)) +
  geom_histogram(binwidth = 10000, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Selling Prices", x = "Selling Price", y = "Frequency")

# Bar plot for seller type
ggplot(vehicle_data, aes(x = seller_type)) +
  geom_bar(fill = "purple") +
  labs(title = "Frequency of Seller Types", x = "Seller Type", y = "Count")

# Bar plot for owner type
ggplot(vehicle_data, aes(x = owner)) +
  geom_bar(fill = "orange") +
  labs(title = "Frequency of Ownership Types", x = "Owner Type", y = "Count")

# Boxplot for kilometers driven
ggplot(vehicle_data, aes(y = km_driven)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Boxplot of Kilometers Driven", y = "Kilometers Driven")

# Scatter plot of kilometers driven vs selling price
ggplot(vehicle_data, aes(x = km_driven, y = selling_price)) +
  geom_point(color = "blue", alpha = 0.6) +
  labs(title = "Selling Price vs Kilometers Driven", x = "Kilometers Driven", y = "Selling Price")

# Boxplot of selling price by seller type
ggplot(vehicle_data, aes(x = seller_type, y = selling_price)) +
  geom_boxplot(fill = "salmon") +
  labs(title = "Selling Price by Seller Type", x = "Seller Type", y = "Selling Price")

# Boxplot of selling price by owner
ggplot(vehicle_data, aes(x = owner, y = selling_price)) +
  geom_boxplot(fill = "cyan") +
  labs(title = "Selling Price by Ownership Type", x = "Owner Type", y = "Selling Price")

# Faceted scatter plot of km driven vs selling price by seller type
ggplot(vehicle_data, aes(x = km_driven, y = selling_price)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ seller_type) +
  labs(title = "Selling Price vs Kilometers Driven by Seller Type", x = "Kilometers Driven", y = "Selling Price")

library(GGally)

# Pairwise plot for selected variables
ggpairs(vehicle_data, columns = c("year", "km_driven", "ex_showroom_price", "selling_price"))
```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values in each column
missing_values <- sapply(vehicle_data, function(x) sum(is.na(x)))
print(missing_values)

# Summary of total missing values in the dataset
total_missing <- sum(missing_values)
cat("Total missing values in the dataset:", total_missing, "\n")

# Display columns with missing values, if any
if (total_missing > 0) {
  cat("Columns with missing values:\n")
  print(missing_values[missing_values > 0])
} else {
  cat("No missing values found in the dataset.\n")
}

```

## Imputation
```{r Imputation}
# Impute missing values in ex_showroom_price with the mean
vehicle_data$ex_showroom_price[is.na(vehicle_data$ex_showroom_price)] <- mean(vehicle_data$ex_showroom_price, na.rm = TRUE)

View(vehicle_data)

# Remove the 'X' column from the dataset
vehicle_data <- vehicle_data[, !(names(vehicle_data) %in% "X")]

# Verify the column has been removed
print(names(vehicle_data))

View(vehicle_data)
```

# Model Training
## Data Splitting
```{r Data Splitting}
library(caret)

# Set seed for reproducibility
set.seed(123)

# Split the data into training (70%) and testing (30%) sets
train_index <- createDataPartition(vehicle_data$selling_price, p = 0.7, list = FALSE)
train_data <- vehicle_data[train_index, ]
test_data <- vehicle_data[-train_index, ]

# Check dimensions of the splits
cat("Training set dimensions:", dim(train_data), "\n")
cat("Testing set dimensions:", dim(test_data), "\n")
```

## Bootstrapping
```{r Bootstrapping}
# Bootstrapping with 100 samples
boot_samples <- 100
boot_results <- list()

for (i in 1:boot_samples) {
  # Resample the training data with replacement
  boot_sample <- train_data[sample(1:nrow(train_data), replace = TRUE), ]
  
  # Fit a model to the bootstrap sample (e.g., linear regression)
  boot_model <- lm(selling_price ~ km_driven + ex_showroom_price, data = boot_sample)
  
  # Store the model results
  boot_results[[i]] <- summary(boot_model)$coefficients
}

# Display results from a sample bootstrapped model
print(boot_results[[1]])
```

## Cross-validation
```{r Cross-validation}
# K-Fold Cross-Validation with 5 folds
train_control <- trainControl(method = "cv", number = 5)
model_cv <- train(selling_price ~ km_driven + ex_showroom_price,
                  data = train_data,
                  method = "lm",
                  trControl = train_control)

# Summary of cross-validation results
print(model_cv)
```

## Training Different Models
```{r Training Different Models}
library(caret)
library(rpart)
library(randomForest)

# Train a Linear Regression model
set.seed(123)
linear_model <- train(selling_price ~ km_driven + ex_showroom_price,
                      data = train_data,
                      method = "lm")

# Display summary of the Linear Regression model
summary(linear_model$finalModel)

# Train a Decision Tree model
set.seed(123)
decision_tree_model <- train(selling_price ~ km_driven + ex_showroom_price,
                             data = train_data,
                             method = "rpart")

# Display the Decision Tree model
print(decision_tree_model$finalModel)

# Plot the decision tree
plot(decision_tree_model$finalModel)
text(decision_tree_model$finalModel, use.n = TRUE)

# Train a Random Forest model
set.seed(123)
random_forest_model <- train(selling_price ~ km_driven + ex_showroom_price,
                             data = train_data,
                             method = "rf",
                             tuneLength = 5)

# Display the Random Forest model
print(random_forest_model)
```

## Performance Comparison
```{r Performance Comparison}
# Make predictions on test data
linear_preds <- predict(linear_model, test_data)
decision_tree_preds <- predict(decision_tree_model, test_data)
random_forest_preds <- predict(random_forest_model, test_data)

# Calculate RMSE for each model
linear_rmse <- RMSE(linear_preds, test_data$selling_price)
decision_tree_rmse <- RMSE(decision_tree_preds, test_data$selling_price)
random_forest_rmse <- RMSE(random_forest_preds, test_data$selling_price)

# Print RMSE results
cat("Linear Regression RMSE:", linear_rmse, "\n")
cat("Decision Tree RMSE:", decision_tree_rmse, "\n")
cat("Random Forest RMSE:", random_forest_rmse, "\n")
```

## Saving Model
```{r Saving Model}
# Load the saved model
loaded_rf_model <- readRDS("./models/saved_random_forest_model.rds")

# New data for prediction (adjust the values based on your input)
new_data <- data.frame(
  km_driven = 15000,
  ex_showroom_price = 100000
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_rf_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

